Analyzing Factors Affecting Pakistan Cricket Team's Performance (1952-2024)

### 2024-12-11 paper/paper.qmd
Updated paper to remove unneccesary words

---
title: "Analyzing Factors Affecting Pakistan Cricket Team's Performance (1952-2024)"
subtitle: "Opponent and Coin Tossplay Play Crucial Role in Deciding Outcome"
author: 
  - Muhammad Abdullah Motasim
thanks: "Code and data are available at: [https://github.com/abdullah-motasim/Analyzing-Pakistans-Cricket-Data](https://github.com/abdullah-motasim/Analyzing-Pakistans-Cricket-Data)."
date: today
date-format: long
abstract: "This paper analyzes Pakistan's cricket performance from 2002 to 2024, using a linear regression model to examine factors influencing match outcomes. We find that opponent strength and toss outcomes are significant predictors of Pakistan's success, highlighting both internal team dynamics and external variables. These findings provide crucial insights for coaches, analysts, and administrators, offering strategies to improve performance under varying conditions. Ultimately, our analysis contributes to a deeper understanding of the factors that drive Pakistan’s cricket success, with implications for teams worldwide."
format: pdf
number-sections: true
bibliography: references.bib
table-of-contents: true
toc-depth: 2
editor: 
  markdown: 
    wrap: 72
---

@@

pakistan_match_data <- read_parquet(here("data/02-analysis_data/cleaned_pakistan_match_data.parquet"))
model <- readRDS(here("models/win_chance.rds"))
confusion_matrix <- readRDS(here("models/confusion_matrix_win_chance.rds"))

```

@@

Pakistan’s cricket team holds a storied legacy, marked by remarkable highs and challenging lows. From their 1992 World Cup triumph to fluctuations in form over the decades, analyzing trends in performance offers insights into their journey and areas for strategic improvement. Despite the wealth of player statistics and match highlights available, there has been a lack of systematic statistical analysis that evaluates long-term performance trends. This paper fills that gap by analyzing data from 2002 to 2024 to uncover the key factors that contribute to Pakistan’s success in international cricket.

We employ a linear regression model to analyze match outcomes,
considering variables like year, opponent, and toss outcomes. Our findings reveal several important insights. First, the strength of the opposing team emerged as a significant factor influencing Pakistan’s chances of winning. We also found that toss outcomes had a noteworthy impact on match results. This paper highlights the importance of both internal factors (such as team performance) and external factors (such as opponent strength) in shaping match outcomes. These insights provide strategic guidance for coaches, analysts, and cricket administrators, not only in Pakistan but also around the world, on how to prepare and perform under varying conditions.

@@

As mentioned before the raw data was collected from the cricketdata
library which contains data on international and other major cricket
matches sourced from ESPN Cricinfo [@espncricinfo] and Cricsheet
[@cricsheet]. It contains details on the individual match year, teams,
result, and toss winner. Additionally, the data set features
comprehensive career statistics for individual players, including total
runs, total balls faced, total wickets, and more. This data set was
chosen for its ease of implementation within R, as it is available as a
package, and for its ability to scrape data from multiple reputable
cricket sources, such as ESPN, thereby enhancing the overall reliability
of the data set.

@@

Figures @fig-match-stats displays the count of several important match
statistics. From these, we can observe that Pakistan faces a variety of
opponents, with the most frequent ones being England, Sri Lanka, and
Australia. Notably, Pakistan has had a challenging record against
Australia, losing a significant number of matches, while many of their
matches ended in draws.

Regarding coin toss outcomes, Sri Lanka and South Africa appear to have
won a substantial number of tosses. However, it's important to note that
while the graphs for match outcomes and tosses won may give the
impression that Pakistan was victorious in most of these events, we must
keep in mind that the data has been filtered to include only Pakistan's
matches. This means Pakistan appears more frequently in the data set
compared to other teams, which accounts for the higher statistics shown.
These figures represent Pakistan's performance specifically, not that of
other teams.

The following two graphs show the number of matches and tosses that
Pakistan won or lost. The win-loss ratio is approximately 0.5,
indicating that Pakistan has lost about twice as many matches as it has
won. Additionally, the coin toss win ratio is close to even, reflecting
a near 50/50 chance of Pakistan winning the toss.

Finally, the last graph illustrates the number of matches Pakistan
played each year. We can see a peak in 2016, with a gradual decline in
matches played over the subsequent years, culminating in a sharp drop in
the most recent year—comparable to the period during the COVID-19
pandemic.

@@

data <- pakistan_match_data %>%
  mutate(pakistan_win = case_when(
    winner == "Pakistan" ~ 1,
    TRUE ~ 0))%>%
  mutate(toss_win = case_when(
    toss_winner == "Pakistan" ~ 1,
    TRUE ~ 0))%>%
  mutate(year = lubridate::year(date))

@@

# Figure 6: Match Results by Year
ggplot(data, aes(x = year)) +
  geom_bar(fill = "brown") +
  labs(x = "Year", y = "Number of Matches") +
  theme_minimal()

```

@@

The model incorporates three independent variables that may influence
match outcomes: year, opponent, and toss winner. The dependent variable
is match_outcome, where 1 indicates a win for Pakistan and 0 indicates a
loss. This setup allows us to examine the influence of key variables on
Pakistan's performance.

By incorporating these variables, the model explores how factors such as
the year of the match, the opposing team, and whether Pakistan won the
toss (which influences the decision to bat or bowl) affect Pakistan’s
likelihood of winning. This approach helps uncover trends in Pakistan's
performance over time and under varying conditions, providing insights
into factors that may inform future match strategies.

@@

-   `Prior for the Coefficients (year, team2, toss_win)`: Each of the
    model coefficients is assigned a Normal prior with a mean of 0 and a
    standard deviation of 2.5. This reflects the assumption that we
    expect the effects of the predictors to be centered around zero,
    with a reasonable amount of uncertainty (a wide spread due to the
    scale of 2.5) about their true values.

-   `Prior for the Intercept`: A normal prior with the same parameters
    as the coefficients (mean = 0, scale = 2.5) is used for the
    intercept. The intercept represents the baseline log-odds of
    Pakistan winning a match when all other predictors are zero.

-   `Prior for the Auxiliary Parameter (prior_aux)`: This parameter
    represents the dispersion of the logistic regression model. An
    Exponential prior with a rate of 1 is used, which is typical in
    Bayesian models for logistic regression. The Exponential prior
    allows for a distribution with only non-negative values, reflecting
    uncertainty in the variance of match outcomes, with higher values
    suggesting more variability in outcomes.

@@

 modelplot(model, coef_omit = "Intercept|team2Ireland") + 
   labs(x = "Coefficient Estimate", y="Predictor Variable")
```

Analyzing some of these coefficients further, The results from the
regression model are as follows:

-   **Intercept**: The intercept of -55.94 (with a standard error of
    80.63) represents the log-odds of Pakistan winning a match when all
    other variables are set to their reference categories. This value is
    quite large in magnitude, but it is not directly interpretable in
    isolation. The negative value suggests a very low baseline
    probability of winning in the absence of other influencing factors.

-   **Year**: The coefficient for year is 0.03 (standard error = 0.04).
    This positive coefficient suggests that, as time progresses, the
    probability of Pakistan winning a match increases very slightly.
    However, the small magnitude of the coefficient indicates that this
    effect is minimal. The relatively narrow standard error suggests
    that the estimate is reasonably precise, but the practical
    significance of this trend is modest.

-   **Opponent**: The coefficients for the opponent variable represent
    how the likelihood of Pakistan winning a match changes when facing
    different teams. These coefficients should be interpreted relative
    to the reference team, which is not explicitly listed here but is
    assumed to be the team with the lowest or most neutral effect:

@@

# Create and save a plot for each opponent
for (opponent in opponents) {

  # Filter the data for the current opponent
  filtered_matches <- pakistan_match_data %>% filter(team2 == opponent)

  # Summarize the number of wins, losses, and draws
  win_loss_summary <- filtered_matches %>%
    group_by(pakistan_result) %>%
    summarise(count = n(), .groups = "drop")
  ggplot(data, aes(x = team2)) +
  geom_bar(fill = "coral") +
  labs(x = "Team", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  # Create the plot
  plot <- ggplot(win_loss_summary, aes(x = pakistan_result, y = count, fill = pakistan_result)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(
      x = "Result",
      y = "Count",
      fill = "Result"
    ) +
    scale_fill_manual(values = c("Win" = "darkgreen", "Loss" = "brown2", "Draw" = "darkgoldenrod2")) +  # Custom colors for Win, Loss, Draw
    coord_cartesian(ylim = c(0, 14)) +
    theme_minimal()+
    theme(legend.position = "none")

@@

Overall, these graphs provide valuable insights into Pakistan's historical performance trends across different opponents. While some teams, such as Australia and England, have posed significant challenges, others, like Bangladesh and Zimbabwe, have been more favorable matchups for Pakistan. These findings align with the regression model coefficients presented earlier, where the opponent variable was a significant predictor of Pakistan's success in matches. For instance, the high win rates against Bangladesh and Zimbabwe correspond to their positive regression coefficients, while the challenging outcomes against teams like Australia reflect negative or smaller coefficients.

@@

pakistan_win_data <- pakistan_match_data %>%
    mutate(pakistan_win = case_when(
    winner == "Pakistan" ~ 1,
    TRUE ~ 0))%>%
  mutate(year = lubridate::year(date))

@@

# Plot individual data points of each player's career runs:
ggplot(filtered_data, aes(x = end, y = bat_runs)) +
  geom_point(color = "darkblue", size = 2) +
  labs(
    # title = "Career Batting Runs for Each Player by Year Career Ended",
    x = "Year Career Ended",
    y = "Career Batting Runs"
  ) +
  theme_minimal()

```

@@

# Plot the aggregated data: Total career runs per year
ggplot(runs_by_year, aes(x = end, y = total_runs)) +
  geom_point(color = "darkblue", size = 3) + 
  labs(
    # title = "Total Career Batting Runs by Year Career Ended",
    x = "Year Career Ended",
    y = "Sum of Total Career Batting Runs"
  ) +
  theme_minimal()

```

@fig-career-end-runs-individual presents a scatter plot showing the years in which players retired and the number of runs they scored throughout their careers, with each dot representing an individual Pakistani player. The plot reveals that most players retired with fewer than 1,000 career runs, and there is generally no strong positive trend across the data. However, there is a notable exception around 2015, where we see a slight uptick in the number of runs. This increase is short-lived, as the trend declines again shortly after, until 2024, when a large number of players retired, many of whom had accumulated over 1,000 career runs. This scatter plot provides valuable insight into individual player performances, which can be used to explore broader trends in the data.

@@

In this paper, I employed Bayesian Logistic Regression to estimate the probability of Pakistan winning cricket matches, considering key factors such as the year of the match, the opponent, and the toss outcome. The model was specifically designed to understand how these variables influence Pakistan's chances of winning, while accounting for uncertainty and variability in the data. To ensure the analysis focused on relevant information, the data was cleaned and processed to include only Pakistan’s matches, along with important player career statistics that could impact match outcomes. By incorporating these factors, the model provided a more comprehensive understanding of Pakistan’s performance dynamics.

The results of the analysis reveal valuable insights into Pakistan's performance trends over time. For example, the strength of the opposing team played a significant role, with match outcomes showing a clear correlation between Pakistan's success and the competitive level of the opposition. Additionally, the toss outcome also emerged as influential factors, providing a deeper understanding of how certain conditions affect Pakistan's likelihood of winning. These findings underscore the importance of both internal (team performance) and external (opponent strength, match context) factors in shaping match outcomes, offering strategic insights for future team preparation and performance analysis.

@@

The model further emphasizes the variations in Pakistan's odds of winning depending on the opponent. For example, the coefficients for teams like Bangladesh, England, and Sri Lanka indicate more favorable matchups for Pakistan, suggesting that these teams have historically been more susceptible to Pakistan's strengths or less competitive overall. Conversely, the coefficients for teams like India and South Africa point to tougher challenges, with Pakistan's chances of winning being lower in these matchups. This pattern highlights the broader strategic landscape, where understanding the relative strength of the opponent becomes crucial for predicting outcomes. It underscores the need for teams to account not just for internal factors, such as player form or match conditions, but also for the competitive level of the opposition when preparing for matches. The results validate the intuition that the quality of the opposing team is a significant determinant of Pakistan’s performance, and that adjusting tactics accordingly can improve chances against stronger teams. Ultimately, the analysis reinforces the importance of a nuanced, strategic approach to match preparation, where the nature of the opposition is carefully considered alongside team strengths and other match variables.

@@

When analyzing @fig-pk-wins in relation to @fig-career-end-runs-individual and @fig-career-end-runs-all, the trends observed align with what we would expect based on the historical performance of Pakistan’s cricket team. From around 2015, we see a noticeable increase in the number of games won by Pakistan. This can be explained by the presence of a number of veteran players who, with their vast experience, played a pivotal role in guiding the team to more victories. Around 2015, several seasoned players reached the peak of their careers or retired, bringing with them not only impressive career statistics but also crucial leadership and expertise on the field. The presence of these experienced players undoubtedly helped stabilize the team's performance and contributed to the higher win rates during this period.

@@

Despite the valuable insights gained, there are several limitations in the data and methodology that must be acknowledged. One significant challenge was the limited number of matches involving Pakistan that could be used to train the model. The data set included only 820 match observations for Pakistan, which, while informative, is relatively small when considering the complexity of cricket as a sport with numerous variables influencing match outcomes. This limited sample size could affect the robustness of the model’s estimates, and in particular, the ability to generalize predictions across different conditions or time periods.

Additionally, there was a lack of detailed match statistics in the raw data, which constrained the scope of the analysis. For instance, while data on match outcomes (win/loss), toss results, and opponents were available, detailed statistics like the number of runs scored, wickets taken, or batting and bowling performances were only provided for the winning team. This lack of comprehensive match data—especially for losing teams—meant that I could not include key performance indicators like runs or wickets in the model, as many of these entries were missing or incomplete. These missing values could have influenced the accuracy of the model, and it would be prudent to consider these missing data points in future iterations of the analysis, potentially employing imputation methods or seeking alternative data sources.

@@

- Improving the Data Set: One of the key areas for future research would be to gather a more extensive data set, particularly focusing on adding detailed match statistics, including runs, wickets, and player performances for both teams. This would allow for a more comprehensive model, potentially including these factors as predictors in a multivariate analysis. Additionally, expanding the sample size to include more matches across a broader time period could help provide more robust and generalizable results.

@@

- Considering Interaction Effects: Future work could explore more advanced models that account for interactions between variables. For instance, the relationship between Pakistan’s performance and the opponent could vary depending on the specific conditions of a given year, such as the emergence of new players, changes in team strategy, or alterations in the competitive landscape of international cricket. Exploring such interaction effects could provide a more nuanced view of Pakistan's performance over time.

@@

## Data Collection
In an ideal scenario, the data would have been far more comprehensive, particularly with respect to the granular details of each match. To overcome the limitations of the current dataset, it would be essential to gather detailed match data through a combination of observational data and video analysis. This would involve either live or recorded footage of each match, from which key statistics could be manually or automatically extracted. Such an approach would include the following steps:

- `Detailed Over-by-Over Data`: Analyzing each over of the match would provide crucial insights into the match's dynamics. This would involve recording key statistics, such as the number of runs scored, the number of wickets taken, the bowler and batter in each over, and any significant events, like boundary hits or wickets. This granular data would help in identifying the ebb and flow of the match and how different events influence Pakistan’s chances of winning.

- `Player-Specific Statistics`: In addition to match-level statistics, it is essential to record individual player data. For every player involved, detailed statistics such as runs scored, wickets taken, the number of innings played, and the number of outs would be recorded. This information would allow for a more precise understanding of how the performance of individual players influences the match outcome. For instance, if certain players consistently score high or take wickets in critical moments, this could provide valuable insight into the team’s overall performance strategy.

@@

## Annual Player Performance
Another critical aspect that could improve the analysis is the collection of annual player performance data. While the current dataset allows for some insight into match outcomes, it lacks detailed player statistics across multiple years. To address this, it would be helpful to build a dataset that combines match-level data for each player over the course of several years.

- `Annual Player Statistics`: For each player, aggregating data across multiple matches would provide a clearer picture of their performance over time. Key metrics such as the total number of runs scored, the number of wickets taken, the number of matches played, and the number of innings would be compiled. This would allow for a more nuanced analysis of individual player performance, considering factors like form and consistency.

@@

- `Player and Opponent Comparisons`: To better understand how player performance interacts with match outcomes, it would be useful to collect data on player performances not just for Pakistan, but also for opposing teams. By comparing the performances of Pakistan’s players with those of the opposing team’s players, we could gain insights into how the relative strength of players affects the result. For example, if Pakistan’s key players perform well but the opposing team’s star players outperform them, it could explain a loss despite good individual performances.

@@

This approach provided a robust evaluation of each model's predictive
performance on unseen data, with the confusion matrix and accuracy
offering insights into the model’s classification capabilities. While
accuracy provides a good indication of overall model performance, we
also recognize that other metrics such as precision, recall, and F1
score might be more informative in the case of imbalanced classes.

@@

# Extract relevant fit diagnostics
fit_diagnostics <- model_summary[c("mean_PPD"),c(1, 3:6)]
fit_diagnostics$variable <- "mean_PPD"
fit_diagnostics <- fit_diagnostics[, c("variable", "mean", "sd", "10%", "50%", "90%")]

@@

fit_diagnostics_rounded |>
  kable(format = "latex")

```

@@

model_summary2 <- summary(model)
model_summary2 <- as.data.frame(model_summary2)
fit_diagnostics2 <- model_summary2[mcmc_row_names,c(2, 7:8)]
fit_diagnostics2$variable <- mcmc_row_names
fit_diagnostics2 <- fit_diagnostics2[, c("variable", "mcse", "Rhat", "n_eff")]

>>>>

---
title: "Analyzing Factors Affecting Pakistan Cricket Team's Performance (1952-2024)"
subtitle: "Opponent and Coin Tossplay Play Key Role in Deciding Outcome"
author: 
  - Muhammad Abdullah Motasim
thanks: "Code and data are available at: [https://github.com/abdullah-motasim/Analyzing-Pakistans-Cricket-Data](https://github.com/abdullah-motasim/Analyzing-Pakistans-Cricket-Data)."
date: today
date-format: long
abstract: "This paper analyzes Pakistan's cricket performance from 2002 to 2024, using a linear regression model to examine factors influencing match outcomes. We find that opponent strength and toss outcomes are significant predictors of Pakistan's success, highlighting both internal team dynamics and external variables. These findings provide information for coaches, analysts, and administrators, offering strategies to improve performance under varying conditions. Ultimately, our analysis contributes to a deeper understanding of the factors that drive Pakistan’s cricket success, with implications for teams worldwide."
format: pdf
number-sections: true
bibliography: references.bib
table-of-contents: true
toc-depth: 2
editor: 
  markdown: 
    wrap: 72
---

@@

pakistan_match_data <- read_parquet(here("data/02-analysis_data/cleaned_pakistan_match_data.parquet"))
model <- readRDS(here("models/win_chance.rds"))
confusion_matrix <- readRDS(here("models/confusion_matrix_win_chance.rds"))
```

@@

Pakistan’s cricket team holds a storied legacy, marked by remarkable highs and challenging lows. From their 1992 World Cup triumph to fluctuations in form over the decades, analyzing trends in performance offers information into their journey and areas for strategic improvement. Despite the wealth of player statistics and match highlights available, there has been a lack of systematic statistical analysis that evaluates long-term performance trends. This paper fills that gap by analyzing data from 2002 to 2024 to uncover the key factors that contribute to Pakistan’s success in international cricket.

We employ a linear regression model to analyze match outcomes,
considering variables like year, opponent, and toss outcomes. Our findings reveal several important observations. First, the strength of the opposing team emerged as a significant factor influencing Pakistan’s chances of winning. We also found that toss outcomes had a noteworthy impact on match results. This paper highlights the importance of both internal factors (such as team performance) and external factors (such as opponent strength) in shaping match outcomes. These observations provide strategic guidance for coaches, analysts, and cricket administrators, not only in Pakistan but also around the world, on how to prepare and perform under varying conditions.

@@

As mentioned before the raw data was collected from the cricketdata
library which contains data on international and other major cricket
matches sourced from ESPN Cricinfo [@espncricinfo] and Cricsheet
[@cricsheet]. It contains details on the individual match year, teams,
result, and toss winner. Additionally, the data set features career statistics for individual players, including total
runs, total balls faced, total wickets, and more. This data set was
chosen for its ease of implementation within R, as it is available as a
package, and for its ability to scrape data from multiple reputable
cricket sources, such as ESPN, thereby enhancing the overall reliability
of the data set.

@@

Figures @fig-match-stats show key match statistics for Pakistan. The data shows that Pakistan frequently faces England, Sri Lanka, and Australia, with a particularly poor record against Australia, including several draws.

Regarding coin toss outcomes, Sri Lanka and South Africa appear to have
won a substantial number of tosses. However, it's important to note that
while the graphs for match outcomes and tosses won may give the
impression that Pakistan was victorious in most of these events, we must
keep in mind that the data has been filtered to include only Pakistan's
matches. This means Pakistan appears more frequently in the data set
compared to other teams, which accounts for the higher statistics shown.

The two graphs showing the number of matches and tosses that
Pakistan won or lost inform us that Pakistan's win-loss ratio is approximately 0.5. This indicates that Pakistan has lost about twice as many matches as it has won. Additionally, the coin toss win ratio is close to even, reflecting
a near 50/50 chance of Pakistan winning the toss.

Finally, the graph illustrates the number of matches Pakistan
played each year has a peak in 2016. It then gradually declines over the subsequent years, culminating in a sharp drop in the most recent year—comparable to the period during the COVID-19 pandemic.

@@

data <- pakistan_match_data %>%
  mutate(pakistan_win = case_when(
    winner == "Pakistan" ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(toss_win = case_when(
    toss_winner == "Pakistan" ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(year = lubridate::year(date))

@@

# Figure 6: Match Results by Year
ggplot(data, aes(x = year)) +
  geom_bar(fill = "brown") +
  labs(x = "Year", y = "Number of Matches") +
  theme_minimal()
```

@@

The model was trained in R utilizing the rstanarm package and it incorporates three independent variables that may influence match outcomes: year, opponent, and toss winner. The dependent variable is match_outcome, where 1 indicates a win for Pakistan and 0 indicates a loss. This setup allows us to examine the influence of key variables on Pakistan's performance.

By incorporating these variables, the model explores how factors such as
the year of the match, the opposing team, and whether Pakistan won the
toss (which influences the decision to bat or bowl) affect Pakistan’s
likelihood of winning. This approach helps uncover trends in Pakistan's
performance over time and under varying conditions, providing intuition
into factors that may inform future match strategies.

@@

-   `Prior for the Coefficients (year, team2, toss_win)`: Each coefficient is assigned a Normal prior with a mean of 0 and a standard deviation of 2.5, reflecting the assumption that the effects of the predictors are centered around zero with a reasonable amount of uncertainty.

-   `Prior for the Intercept`: A Normal prior with a mean of 0 and standard deviation of 2.5 is used, representing the baseline log-odds of Pakistan winning a match when all predictors are zero.

-   `Prior for the Auxiliary Parameter (prior_aux)`: This represents the dispersion of the logistic regression model and is assigned an Exponential prior with a rate of 1, typical for Bayesian logistic regression. It reflects uncertainty in outcome variance, with higher values suggesting more variability.

@@

modelplot(model, coef_omit = "Intercept|team2Ireland") +
  labs(x = "Coefficient Estimate", y = "Predictor Variable")
```

The regression model results are as follows:

-   **Intercept**: The intercept of -55.94 (standard error = 80.63) represents the log-odds of Pakistan winning when all other variables are at their reference categories. While large in magnitude, it is not directly interpretable. The negative value suggests a very low baseline probability of winning without other influencing factors.

-   **Year**: The coefficient for year is 0.03 (standard error = 0.04), indicating a very slight increase in the probability of Pakistan winning over time. The small magnitude and narrow standard error suggest the effect is minimal and the estimate is fairly precise.

-   **Opponent**: The coefficients for opponents show how Pakistan’s likelihood of winning changes against different teams, relative to the reference team:

@@

# Create and save a plot for each opponent
for (opponent in opponents) {
  # Filter the data for the current opponent
  filtered_matches <- pakistan_match_data %>% filter(team2 == opponent)

  # Summarize the number of wins, losses, and draws
  win_loss_summary <- filtered_matches %>%
    group_by(pakistan_result) %>%
    summarise(count = n(), .groups = "drop")
  ggplot(data, aes(x = team2)) +
    geom_bar(fill = "coral") +
    labs(x = "Team", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  # Create the plot
  plot <- ggplot(win_loss_summary, aes(x = pakistan_result, y = count, fill = pakistan_result)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(
      x = "Result",
      y = "Count",
      fill = "Result"
    ) +
    scale_fill_manual(values = c("Win" = "darkgreen", "Loss" = "brown2", "Draw" = "darkgoldenrod2")) + # Custom colors for Win, Loss, Draw
    coord_cartesian(ylim = c(0, 14)) +
    theme_minimal() +
    theme(legend.position = "none")

@@

Overall, these graphs provide useful information into Pakistan's historical performance trends across different opponents. While some teams, such as Australia and England, have posed significant challenges, others, like Bangladesh and Zimbabwe, have been more favorable matchups for Pakistan. These findings align with the regression model coefficients presented earlier, where the opponent variable was a significant predictor of Pakistan's success in matches. For instance, the high win rates against Bangladesh and Zimbabwe correspond to their positive regression coefficients, while the challenging outcomes against teams like Australia reflect negative or smaller coefficients.

@@

pakistan_win_data <- pakistan_match_data %>%
  mutate(pakistan_win = case_when(
    winner == "Pakistan" ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(year = lubridate::year(date))

@@

# Plot individual data points of each player's career runs:
ggplot(filtered_data, aes(x = end, y = bat_runs)) +
  geom_point(color = "darkblue", size = 2) +
  labs(
    # title = "Career Batting Runs for Each Player by Year Career Ended",
    x = "Year Career Ended",
    y = "Career Batting Runs"
  ) +
  theme_minimal()
```

@@

# Plot the aggregated data: Total career runs per year
ggplot(runs_by_year, aes(x = end, y = total_runs)) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    # title = "Total Career Batting Runs by Year Career Ended",
    x = "Year Career Ended",
    y = "Sum of Total Career Batting Runs"
  ) +
  theme_minimal()
```

@fig-career-end-runs-individual presents a scatter plot showing the years in which players retired and the number of runs they scored throughout their careers, with each dot representing an individual Pakistani player. The plot shows that most players retired with fewer than 1,000 career runs, and there is generally no strong positive trend across the data. However, there is a notable exception around 2015, where we see a slight uptick in the number of runs. This increase is short-lived, as the trend declines again shortly after, until 2024, when a large number of players retired, many of whom had accumulated over 1,000 career runs. This scatter plot provides useful information on individual player performances, which can be used to explore broader trends in the data.

@@

In this paper, I employed Bayesian Logistic Regression to estimate the probability of Pakistan winning cricket matches, considering key factors such as the year of the match, the opponent, and the toss outcome. The model was specifically designed to understand how these variables influence Pakistan's chances of winning, while accounting for uncertainty and variability in the data. To ensure the analysis focused on relevant information, the data was cleaned and processed to include only Pakistan’s matches, along with important player career statistics that could impact match outcomes. By incorporating these factors, the model provided a more complete understanding of Pakistan’s performance dynamics.

The results of the analysis reveal useful intuitions into Pakistan's performance trends over time. For example, the strength of the opposing team played a significant role, with match outcomes showing a clear correlation between Pakistan's success and the competitive level of the opposition. Additionally, the toss outcome also emerged as influential factors, providing a deeper understanding of how certain conditions affect Pakistan's likelihood of winning. These findings underscore the importance of both internal (team performance) and external (opponent strength, match context) factors in shaping match outcomes, offering strategic information for future team preparation and performance analysis.

@@

The model further emphasizes the variations in Pakistan's odds of winning depending on the opponent. For example, the coefficients for teams like Bangladesh, England, and Sri Lanka indicate more favorable matchups for Pakistan, suggesting that these teams have historically been more susceptible to Pakistan's strengths or less competitive overall. Conversely, the coefficients for teams like India and South Africa point to tougher challenges, with Pakistan's chances of winning being lower in these matchups. This pattern highlights the broader strategic landscape, where understanding the relative strength of the opponent becomes important for predicting outcomes. It underscores the need for teams to account not just for internal factors, such as player form or match conditions, but also for the competitive level of the opposition when preparing for matches. The results validate the intuition that the quality of the opposing team is a significant determinant of Pakistan’s performance, and that adjusting tactics accordingly can improve chances against stronger teams. Ultimately, the analysis reinforces the importance of a strategic approach to match preparation, where the nature of the opposition is carefully considered alongside team strengths and other match variables.

@@

When analyzing @fig-pk-wins in relation to @fig-career-end-runs-individual and @fig-career-end-runs-all, the trends observed align with what we would expect based on the historical performance of Pakistan’s cricket team. From around 2015, we see a noticeable increase in the number of games won by Pakistan. This can be explained by the presence of a number of veteran players who, with their vast experience, played a pivotal role in guiding the team to more victories. Around 2015, several seasoned players reached the peak of their careers or retired, bringing with them not only impressive career statistics but also leadership and expertise on the field. The presence of these experienced players undoubtedly helped stabilize the team's performance and contributed to the higher win rates during this period.

@@

Despite the useful information gained, there are several limitations in the data and methodology that must be acknowledged. One significant challenge was the limited number of matches involving Pakistan that could be used to train the model. The data set included only 820 match observations for Pakistan, which, while informative, is relatively small when considering the complexity of cricket as a sport with numerous variables influencing match outcomes. This limited sample size could affect the robustness of the model’s estimates, and in particular, the ability to generalize predictions across different conditions or time periods.

Additionally, there was a lack of detailed match statistics in the raw data, which constrained the scope of the analysis. For instance, while data on match outcomes (win/loss), toss results, and opponents were available, detailed statistics like the number of runs scored, wickets taken, or batting and bowling performances were only provided for the winning team. This lack of extensive match data—especially for losing teams—meant that I could not include key performance indicators like runs or wickets in the model, as many of these entries were missing or incomplete. These missing values could have influenced the accuracy of the model, and it would be prudent to consider these missing data points in future iterations of the analysis, potentially employing imputation methods or seeking alternative data sources.

@@

- Improving the Data Set: One of the key areas for future research would be to gather a more extensive data set, particularly focusing on adding detailed match statistics, including runs, wickets, and player performances for both teams. This would allow for a more sophisticated model, potentially including these factors as predictors in a multivariate analysis. Additionally, expanding the sample size to include more matches across a broader time period could help provide more robust and generalizable results.

@@

- Considering Interaction Effects: Future work could explore more sophisticated models that account for interactions between variables. For instance, the relationship between Pakistan’s performance and the opponent could vary depending on the specific conditions of a given year, such as the emergence of new players, changes in team strategy, or alterations in the competitive landscape of international cricket. Exploring such interaction effects could provide a more refined view of Pakistan's performance over time.

@@

## Data Collection
In an ideal scenario, the data would have been far more extensive, particularly with respect to the granular details of each match. To overcome the limitations of the current dataset, it would be essential to gather detailed match data through a combination of observational data and video analysis. This would involve either live or recorded footage of each match, from which key statistics could be manually or automatically extracted. Such an approach would include the following steps:

- `Detailed Over-by-Over Data`: Analyzing each over of the match would provide important data into the match's dynamics. This would involve recording key statistics, such as the number of runs scored, the number of wickets taken, the bowler and batter in each over, and any significant events, like boundary hits or wickets. This granular data would help in identifying the ebb and flow of the match and how different events influence Pakistan’s chances of winning.

- `Player-Specific Statistics`: In addition to match-level statistics, it is essential to record individual player data. For every player involved, detailed statistics such as runs scored, wickets taken, the number of innings played, and the number of outs would be recorded. This information would allow for a more precise understanding of how the performance of individual players influences the match outcome. For instance, if certain players consistently score high or take wickets in key moments, this could provide useful information into the team’s overall performance strategy.

@@

## Annual Player Performance
Another important aspect that could improve the analysis is the collection of annual player performance data. While the current dataset allows for some intuition into match outcomes, it lacks detailed player statistics across multiple years. To address this, it would be helpful to build a dataset that combines match-level data for each player over the course of several years.

- `Annual Player Statistics`: For each player, aggregating data across multiple matches would provide a clearer picture of their performance over time. Key metrics such as the total number of runs scored, the number of wickets taken, the number of matches played, and the number of innings would be compiled. This would allow for a more refined analysis of individual player performance, considering factors like form and consistency.

@@

- `Player and Opponent Comparisons`: To better understand how player performance interacts with match outcomes, it would be useful to collect data on player performances not just for Pakistan, but also for opposing teams. By comparing the performances of Pakistan’s players with those of the opposing team’s players, we could gain knowledge on how the relative strength of players affects the result. For example, if Pakistan’s key players perform well but the opposing team’s star players outperform them, it could explain a loss despite good individual performances.

## Simulation Study
A key aspect of ensuring the robustness of the data is to evaluate the impact of potential sampling bias on performance predictions. By running a simulation study, we could artificially create datasets with varying levels of completeness and bias (e.g., underrepresentation of certain match conditions or players) and test how these biases affect the ability to predict match outcomes. For instance, simulations could explore scenarios where certain conditions, such as playing at home or under specific weather conditions, are underrepresented in the data. By comparing the outcomes of these simulations with real-world performance predictions, we can better understand how bias in sampling influences model accuracy and the generalizability of results.

## Bias-Adjustment Techniques
In addition to simulation, statistical techniques such as propensity score matching or inverse probability weighting could be applied to adjust for biases in the observed data. These methods would help to correct for any over- or under-representation of certain conditions and ensure that the analysis is reflective of a broader set of match scenarios. For example, if the dataset is skewed towards games played against stronger teams, these techniques could help balance the data to ensure that the analysis is not overly influenced by these high-stakes matchups.

@@

This approach provided a robust evaluation of each model's predictive
performance on unseen data, with the confusion matrix and accuracy
offering information on the model’s classification capabilities. While
accuracy provides a good indication of overall model performance, we
also recognize that other metrics such as precision, recall, and F1
score might be more informative in the case of imbalanced classes.

@@

# Extract relevant fit diagnostics
fit_diagnostics <- model_summary[c("mean_PPD"), c(1, 3:6)]
fit_diagnostics$variable <- "mean_PPD"
fit_diagnostics <- fit_diagnostics[, c("variable", "mean", "sd", "10%", "50%", "90%")]

@@

fit_diagnostics_rounded |>
  kable(format = "latex")
```

@@

model_summary2 <- summary(model)
model_summary2 <- as.data.frame(model_summary2)
fit_diagnostics2 <- model_summary2[mcmc_row_names, c(2, 7:8)]
fit_diagnostics2$variable <- mcmc_row_names
fit_diagnostics2 <- fit_diagnostics2[, c("variable", "mcse", "Rhat", "n_eff")]
